---
title: "Shortest Path Optimization in R"
categories:
  - network flows
  - linear programming
  - r
description: |
  In this post we will walk through how to make least cost decisions using network flows shortest path algorithm.
author:
  - name: Blake Conrad
    url: https://github.com/conradbm
date: 2022-07-09
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
---

```{r setup, echo=FALSE, include=F}
library(dplyr)
library(igraph)
library(visNetwork)
library(rmarkdown)
```
 
# Vehicle Replacement Using Shortest Path 

A common problem we often face in the world is to find the shortest path between two points. Whether it is on the road, or how to obtain an object in in our intellectual trajectory, we are always seeking to optimize. 

In network flows one common problem is to find the `s-t shortest path`. The problem formulation is as follows.

Given some net of nodes, and two particular nodes `s` and `t` that are not the same, find the shortest distance through the edges connecting them. The mathematical formulation is below.
$$ N \in \{n_1, n_2, .., n_D\} $$
$$ E = \{N \times N\} = \{e_{ij} \in \Re^+ | \:\:\forall i,j \in N \} $$
$$ s,t \in N : s\neq t$$


## Problem Setup

We all have to drive cars at some point in our adult lives (well, most of us). With this comes the question of <em>investment decisions</em> if I am allowed to call it that. Which car should we buy? How long should we keep it? Is it more prudent to keep paying maintenance, repairs, and damages, or just get a new ride?

We would like to know what is the best decision to make over the next 5 years for our vehicle needs. So we take this problem and model it as an optimization problem using the famous `shortest paths algorithm`.

So with our problem the decision space is pretty straight forward:

1. Each year we can choose to keep our vehicle, or

2. Trade it in

However, every year we choose to keep our vehicle, we must pay maintenance costs for it. So every year we keep it, there is a cumulative maintenance cost. Once we trade-in we offset the cost of the new car with the trade in value, and pay much less maintenance on the new ride. Let's take a look at the problem data.

## Problem Data

We know the cost every year of a new vehicle is assumed as `$12,000` for simplicity. Further, we have some records of what maintenance costs and trade-in values will be.

### New Car Cost
The new car cost is assumed constant every year. An interesting homework assignment would be to make this `stochastic` and change over time. This is much more suitable to the real world, but for this example will remain constant.

```{r new_car_cost}
NEW_CAR_COST <- 12000
NEW_CAR_COST
```


### Maintenance Costs
The maintenance costs are for 4 years. Each year you keep the car, you will pay more on maintenance.

```{r maintenance_cost}
M <- data.frame(Year = c(0:4), Maintenance_cost = c(2000, 4000, 5000, 9000, 12000))
M_vec <- M %>% dplyr::pull(Maintenance_cost)
M
```

### Trade-in Revenues
The trade-in price is similar to maintenance. Each year you keep the vehicle, it will depreciate. So to account for this we have a decreasing trade-in value.

```{r trade_in_revenue}
T <- data.frame(Year = c(1:5), Trade_in_price = c(7000, 6000, 2000, 1000, 0))
T_vec <- T %>% dplyr::pull(Trade_in_price)
T
```

### Cost Matrix
Since we know the costs will be cumulative, so we know what each years will be. The cost matrix will be for the number of years the car is kept to accumulate costs from maintenance. In mathematical language, this is represented below:

$$ c_{ij} = \sum_{t=1}^{j-1}{M_{t-1}}\:\: if\: j > i \:\: otherwise \: \infty $$

Where `M` is the maintenance matrix defined above. 

We also know our objective is to minimize the total cost, which equates to `maintanence cost` + `cost to purchase a new car` - `trade in value`.

```{r cost_matrix}
# Nodes dataframe
nodes = data.frame(Year = c(sprintf("Year %s", seq(1:6))),
                   Color = c("green", "gold", "gold", "gold", "gold", "red"))
n = nrow(nodes)

# Edges list
edges = list(from=c(), to=c(), cost=c(), color=c())

# Cost matrix
C <- matrix(0, n, n)
BIG_M <- 1000000
for(i in 1:n){
  for (j in 1:n){
    if(j > i){
      
      # Cost of maintenance
      maintenance_cost <- M_vec[1:(j-i)]
      maintenance_cost_total <- sum(maintenance_cost)

      # Cost of new car
      new_car_cost = NEW_CAR_COST
      
      # Trade-in value
      trade_in_revenue <- T_vec[j-i]

      # Total cost for decision to buy car on year i and sell it on year j
      total_cost_for_decision_i_to_j <- new_car_cost + maintenance_cost_total - trade_in_revenue

      # Save the value into cost matrix
      C[i,j] <- total_cost_for_decision_i_to_j
      edges$from <- append(edges$from, paste("Year", i))
      edges$to <- append(edges$to, paste("Year", j))
      edges$cost <- append(edges$cost, total_cost_for_decision_i_to_j)
      edges$color <- append(edges$color, "grey")
    }
    else{
      
      # Big M otherwise to make edge infeasible
      C[i,j] <- BIG_M
    }
  }
}

# Edges dataframe
edges <- edges %>% as.data.frame

nodes
edges
C
```


## Solving the Problem

Now that we have our cost matrix, the last ingredient is to solve the problem. That means to solve the `s-t shortest path` from `Year 0` to `Year 6`, so we can determine what is the cheapest investment strategy for us. For this, we will be using the `igraph` package.


### Visualizing the network

```{r viz_network}

g <- igraph::graph_from_data_frame(d = edges,
                                   directed = TRUE,
                                   vertices = nodes)
# Make it a little cleaner
plot(g,
     main = "Cost of Annual Vehicle Trade-in",
     edge.arrow.size=.5,
     vertex.color=V(g)$Color,
     edge.color=E(g)$color,
     edge.label = E(g)$cost,
     edge.width = E(g)$cost/10000,
     vertex.size=20,
     vertex.frame.color="gray",
     vertex.label.color="black",
     vertex.label.cex=0.8,
     vertex.label.dist=2, layout = layout_as_star,
     edge.curved=0.2)

```

### Shortest Path Solution

So once we perform the `Dijkstra's Shortest Path` algorithm on the network we obtain a solution in matrix form. This solution tells us the best possible cost for our car decision from `Year 1` to `Year 6` will be at **$31,000**.

What this does not show us is what path was chosen to obtain that value.

```{r solve_sp}
# Get the shortest paht cost matrix
s_paths_cost <- igraph::shortest.paths(graph = g, v = V(g), weights = E(g)$cost, algorithm = "dijkstra")
s_paths_cost

```

### Shortest Path Analysis

The optimal selection is the following sequence:

- Buy a new car in `Year 1` keep the car for a year, then sell on `Year 2`.
- Buy another car in `Year 2`, but keep for two years and sell on `Year 4`.
- Finally, buy a car on `Year 4`, keep for two years and sell on `Year 6`.

Do these numbers add up? Let's check. 7000 + 12000 + 12000 == 31000 is `r 7000 + 12000 + 12000 == 31000`.

So our least cost strategy can be no less than `$31000` over the next 6 years. With the current cost structure, means to keep the car for a year or two, then pitch it because the trade-off between maintenance accumulation and depreciation start to mutually deter from a least cost decision.

```{r sp_analysis}

# Get all path distances solution vertex path
s.paths <- igraph::shortest_paths(graph = g,
                                  from = "Year 1",
                                  output = "vpath",
                                  weights = E(g)$cost, 
                                  to = "Year 6")
                                  # v = V(g), 
                                  # to = V(g), 
                                  # weights = E(g)$cost)

# Update colors from vertex path found
s.paths$vpath
E(g, path = s.paths$vpath[[1]])$color <- "red"
```

```{r final_plot}
plot(g,
     main = "Least Cost Vehicle Trade-in Policy",
     edge.arrow.size=.5,
     vertex.color=V(g)$Color,
     edge.color=E(g)$color,
     edge.label = E(g)$cost,
     edge.width = E(g)$cost/10000,
     vertex.size=20,
     vertex.frame.color="gray",
     vertex.label.color="black",
     vertex.label.cex=0.8,
     vertex.label.dist=2, layout = layout_as_star,
     edge.curved=0.2)
```

This plot doesn't look too great! Let's try to spruce it up a bit using `visNetwork`, a common package in R that leverages the `vis.js` framework, which can be found [here](https://visjs.org/).

```{r visNetwork_attempt}
V(g)$label = nodes$Year
V(g)$shape = "circle"
E(g)$width = edges$cost/10000
E(g)$weight = edges$cost/10000
E(g)$label = edges$cost %>% as.character

visIgraph(igraph = g)
```

# References

Example taken from the following sources:

<cite>Winston., Wayne. Operations Research, Applications and Algorithms 4th Edition.</cite>

