---
title: "Transportation Problem in R"
description: |
  A short description of the post.
categories:
  - network flows
  - linear programming
author:
  - name: Blake Conrad
    url: https://github.com/conradbm
date: 2022-07-12
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
library(lpSolve)
library(igraph)
library(visNetwork)
library(dplyr)
library(DiagrammeR)
library(ggplot2)
```

# Metropolitan Power Plant Supply

## The Problem

### Description

One of the most common daily tasks is that of transportation. From our bed to the dining room table, or from home to work and back, all of these examples have limited **supply** some pre-determined **demand** and an implicit **cost** in order to carry it out. The question is, what is the best way to do it? It would be silly to travel to go home before you pick up dinner, then go back out again to do it; we seek to **minimize our cost** when transporting goods. This can be time, money, risk, or anything else that matters. 

In our example we will consider a power plant that needs to supply electricity (or any resource for that matter) to a city. All of the costs incurred to transport the electricity are known, these will be in USD. Supply `from` each power plant is known. Demand `to` each city is also known. The problem is 

### Mathematical Formulation

The classical linear programming formulation goes as follows:

*Objective*

$$
Minimize. \sum_{i,j}{x_{ij}c_{ij}}
$$

$$
s.t.
$$

*Supply Constraints*

$$
\sum_{j}{x_{ij}} \le s_i \: \forall i \in S
$$

*Demand Constraints*

$$
\sum_{i}{x_{ij}} \ge d_j \: \forall j \in D
$$

*Non-negativity Constraints*

$$
x_{ij} \ge 0 \: \forall (i,j)\in A
$$

### Data

First we need to get our data. This will come in the form of an `adjacency matrix` which we will readily convert into an arc matrix for reasons that will become clear soon (building constraints for the LP).

```{r adj_matrix}
#Number of plants and cities
NUM_POWER_PLANTS = 3
NUM_CITIES = 4

# Adjacency matrix
adj.matrix = matrix(c(8, 6, 10, 9, 9, 12, 13, 7, 14, 9, 16, 5), 
                    nrow = NUM_POWER_PLANTS, 
                    ncol = NUM_CITIES)
adj.matrix
```

Next we need to take a look at our supply and demand.

```{r supply_demand}

# Supply and demand vectors
supply <- c(35, 50, 40)
demand <- c(45, 20, 30, 30)
supply
demand
```

Next define the Power Plants and Cities. 

```{r names}

# Supply node id lookup
POWER_PLANTS = 1:3
POWER_PLANT_LABELS <- sprintf("Plant %s", 1:3)

# Demand node id lookup
CITIES = 1:4
CITIES_LABELS <- sprintf("City %s", 1:4)

POWER_PLANT_LABELS
CITIES_LABELS
```


For the nodes we will supply them with some metatdata into the `DiagrammeR` package for slick plotting capability.

```{r nodes}

# Create the powerplant nodes
powerplant_nodes <- DiagrammeR::create_node_df(nodes = POWER_PLANTS,
                                  label = POWER_PLANT_LABELS,
                                  style = "filled",
                                  type = "upper",
                                  color = rep("green", length(POWER_PLANTS)),
                                  shape = rep("circle", length(POWER_PLANTS)),
                                  data = supply,
                                  n = length(POWER_PLANTS)
                                  )

# Create the city nodes
city_nodes <- DiagrammeR::create_node_df(nodes = CITIES,
                                  label = CITIES_LABELS,
                                  style = "filled",
                                  type = "lower",
                                  color = rep("red", length(CITIES)),
                                  shape = rep("square", length(CITIES)),
                                  data = demand,
                                  n = length(CITIES)
                                  )

# Create the DiagrammeR dataframe
nodes <- DiagrammeR::combine_ndfs(powerplant_nodes, city_nodes)
nodes
```

One way to convert the `adjacency matrix` into a friendlier edge format is to simply `gather` it with some basic naming of the rows and columns.
Once we have the `tidy_edges` we can acquire the `node_id` for each by doing a little joining with the `nodes` table formed above. This will supply the `id's` required for the `DiagrammeR` package.

```{r edges}

# Create the "from" "to" representation, from the raw matrix form
tmp <- adj.matrix %>% as.data.frame
names(tmp) <- CITIES_LABELS
tmp$from <- POWER_PLANT_LABELS
tidy_edges <- tmp %>% 
                tidyr::gather(., key = "to", value = "data", -c(from)) %>%
                dplyr::mutate(color = "black", rel = "requires")

# Go find the from_id
from_id = tidy_edges %>% 
            dplyr::inner_join(nodes, by = c("from"="label")) %>% 
            dplyr::transmute(from_id = id) %>%
            dplyr::pull(from_id)

# Go find the to_id
to_id = tidy_edges %>% 
            dplyr::inner_join(nodes, by = c("to"="label")) %>% 
            dplyr::transmute(to_id = id) %>% 
            dplyr::pull(to_id)


# Get the from_id
tidy_edges$from_id = from_id

# Get the to_id
tidy_edges$to_id = to_id

edges <- DiagrammeR::create_edge_df(from = from_id,
                                    to = to_id,
                                    rel = tidy_edges$rel,
                                    color = tidy_edges$color,
                                    data = tidy_edges$data,
                                    label = tidy_edges$data)


# Always a good idea to short your edges logically
edges <- edges %>% 
            dplyr::arrange(-desc(from), -desc(to)) %>% 
            dplyr::mutate(id = row_number())
edges
```


### A Quick Visual

```{r create_vis}
# Create the graph
g = DiagrammeR::create_graph(nodes_df = nodes,
                             edges_df = edges, 
                             directed = T, 
                             graph_name = "transportation")

# Render the graph
DiagrammeR::render_graph(g, title = "Power Plants Electricity Transportation to Cities")
```

Another way to visualize this is by converting it to an igraph, which looks pretty horrible.

```{r}
ig <- DiagrammeR::to_igraph(g)
plot(ig, vertex.size=30)
```


# LP Model Formulation

## Objective

First things first, we know the costs on each arc already, because they are supplied from our `adjacency matrix` initially. So we can just go recover those.

```{r objective}
# Get objective value
f.obj = edges %>% dplyr::pull(data)
f.obj
```

## Constraints, Direction, and RHS

First some convenient numbering systems will be helpful to preserve knowledge of the nodes and which id is which.

```{r node_ids}
# Align ids up for matrix building
POWER_PLANT_NODE_IDS <- POWER_PLANTS
CITY_NODE_IDS <- POWER_PLANTS[length(POWER_PLANTS)] + CITIES

POWER_PLANT_NODE_IDS
CITY_NODE_IDS
```


Next we need to build the supply constraints. So for each unique powerplant arc, then sum of it's outbound arcs must be less than the supply constraint, for all supply.

```{r supply_constraints}

# One constraint row for each supply node
Smat = matrix(0, nrow = length(POWER_PLANTS), ncol = nrow(edges))
Smat_rhs = supply
Smat_dir = rep("<=", length(supply))

for(i in 1:nrow(Smat)){
  for(j in 1:ncol(Smat)){
    from = edges$from[j]
    to = edges$to[j]
    
    if(from == POWER_PLANT_NODE_IDS[i]){
      Smat[i, j] = 1
    }
  }
}

Smat
Smat_rhs
Smat_dir
```

With the same logic we need to construct the demand matrix.

```{r}

# One constraint node for each demand node
Dmat = matrix(0, nrow=length(CITIES), ncol = nrow(edges))
Dmat_rhs = demand
Dmat_dir = rep(">=", length(demand))

for(i in 1:nrow(Dmat)){
  for(j in 1:ncol(Dmat)){
    from = edges$from[j]
    to = edges$to[j]
    
    if(to == CITY_NODE_IDS[i]){
      Dmat[i, j] = 1
    }
  }
}

Dmat
Dmat_rhs
Dmat_dir
```


Now that we have our supply and demand data matrices put together, let's unify them and take a look at all of our constraints.

```{r}

# All `lpSolve` data
f.obj <- f.obj
f.cons <- rbind(Smat, Dmat)
f.rhs <- c(Smat_rhs, Dmat_rhs)
f.dir <- c(Smat_dir, Dmat_dir)

f.obj
f.cons
f.rhs
f.dir
```


## Solution

Now we unite everything together and drop it into the solver. Remember, we are trying to minimize cost in our objective, the constraints will maximize the flow. So specify `min`.

```{r solve_lp}

# Get the results
results <- lp(direction = "min",  
              objective.in = f.obj, 
              const.mat = f.cons, 
              const.dir = f.dir, 
              const.rhs = f.rhs)

results$objval
matrix(results$solution, nrow = length(supply), ncol = length(demand))
```

### Visualization

Cool! We have our solution. It looks like:

- `Plant 1` should supply `City 1` and `City 2`.
- `Plant 2` should supply `City 1` and `City 3`, and
- `Plant 3` should supply `City 1` and `City 4`.

One interesting insight is the shared responsibility all plants have in getting resources to `City 1`.
Another insight is the partnerships to make `City 1` supplied as cheaply as possible, may require some additional `coordination costs` to carefully handle any error in costs between ill-shipments.
The last insights are the siloe's, `City 2` is entirely handled by `Plant 1`, `City 3` by `Plant 2`, and `City 4` by `Plant 3`. This shows us the reliance we have on those energy transfers going through, otherwise it's not only lights out, it is also a more expensive transfer to get them back on!

```{r final_visual}

# Add in the flow
edges$flow <- results$solution

# Color things if there exists flow
edges <- edges %>% dplyr::mutate(weight = flow,
                                 label = flow,
                                 color = if_else(condition = flow > 0, true = "black",false = "grey"))

# Create the graph
g = DiagrammeR::create_graph(nodes_df = nodes,
                             edges_df = edges, 
                             directed = T, 
                             graph_name = "transportation")

# Draw the graph
DiagrammeR::render_graph(g, title = "Power Plants Electricity Transportation to Cities")
```

```{r}
# ig <- DiagrammeR::to_igraph(g)

nodes
visNet_nodes <- nodes %>% 
                dplyr::mutate(shape = "circle",
                              font = 12,
                              size = data,)
edges
visNet_edges <- edges %>% 
                dplyr::mutate(label = paste("(",flow,",",data,")",sep=""),
                              length = 250,
                              width = flow/5,
                              arrows = "to") 
  
visNetwork(visNet_nodes, 
           visNet_edges,
           width = "100%",
           main = "Power Distribution | (flow, cost)")
```


# An Easier Way

It's almost spooky how easy it is with the `lp.transport` API. Verify that we got the same solution, `880` and the decision vector is also the same. Since the objective value is the same, the solutions might differ, because they are equally optimal. Pretty interesting to see how easy it is with the convenience of `lpSolve` though!

```{r}

# Get the results
results <- lpSolve::lp.transport(cost.mat = adj.matrix, 
                                  direction = "min", 
                                  row.signs = rep("<=", nrow(adj.matrix)), 
                                  col.signs = rep(">=", ncol(adj.matrix)),
                                  row.rhs = supply, 
                                  col.rhs = demand)

# Verify `objval` is the same as above
results$objval
results$solution
```


