---
title: "Maximum Network Flows in R"
categories:
  - network flows
  - linear programming
  - r
description: |
  In this post we will walk through how to make a maximum flow decision using network flows and linear programming.
author:
  - name: Blake Conrad
    url: https://github.com/conradbm
date: 2022-07-09
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
---


```{r packages, include=FALSE}
library(dplyr)
library(lpSolve)
library(igraph)
library(visNetwork)
library(rmarkdown)
```

# Oil Flow Maximization

One classic problem in `Network Flows` and `Optimization` is called the `Max-Flow Problem`. This takes any two nodes in a network, `s` and `t` and attempts to send as much of a resource (or multiple) from `s` to `t`. This is called a `flow`, and the flow which maximizes the total bandwidth of the network is called the `maximum flow`.

First, the problem starts with an objective: to `maximize flow`. These are denoted as,

$$x_{ij} = flow \: on \: node_i \: to \: node_j \:(or \: on \: arc \: (i,j))$$ 


Second, the problem has a set of constraints, these are called the `arc capacities`. These are denoted as,

$$u_{ij} = maximum \:amount \:of \: feasible \: flow \: on \: node_i \: to \: node_j \:(or \: on \: arc \: (i,j))$$ 
Last, the network graph is supplied as a set of connections under the traditional network structure:

$$ G = (N,E) $$

# Problem Formation

For our problem, the feasible flow is going to be in units of millions of barrels of oil per hour that will pass through an arc of pipeline.

## Network Structure

The source for our network is indicated in green and sink in red.

```{r data_setup}
nodes <- data.frame(id = c(0:4), color = c("green", rep("grey", 3), "red"))
edges <- data.frame(from = c(0,0,1,1,3,2), to = c(1,2,2,3,4,4), capacity = c(2,3,3,4,1,2), color = "grey")

edges
```
## Network Visual

```{r initial_graph}
g <- igraph::graph_from_data_frame(d = edges,
                                   directed = TRUE,
                                   vertices = nodes)
plot(g, edge.label = E(g)$capacity, 
     main = "Oil Pipeline Capacities", 
     vertex.color = V(g)$color,
     edge.color = E(g)$color)
```

## Big M

In order to model this problem, we need to put a very large capacity from the `source node` to the `sink node`, these are `s` and `t` mentioned above. We have given them node names `0` and `4` respectively in the dataframe for `nodes`.

```{r updated_graph}
BIG_M <- 1000000

edges <- rbind(edges, data.frame(from = c(4),
                                 to = c(0), 
                                 capacity = c(BIG_M), 
                                 color = "purple"))

# Never hurts to add an id
edges$id <- 0:(nrow(edges)-1)

g <- igraph::graph_from_data_frame(d = edges,
                                   directed = TRUE,
                                   vertices = nodes)
plot(g,
     edge.label = E(g)$capacity, 
     main = "Oil Pipeline Capacities", 
     vertex.color = V(g)$color,
     edge.color = E(g)$color)
```

# Linear Programming (LP)

## LP Structure

The next step is to set up the optimization. Let's do that now. There are 3 key ingredients.

- Objective

- Constraints

- Directions

### Constraints

We want to build the constraint matrix, which has 2 parts.

- Arc Capacity Constraints

- Node Flow Constraints

The `Arc Capacity Constraints` are the first we address.

The `Amat`, or A matrix, is the arc matrix which contains the upper bounds. Since linear programming relies on the resource matrix, we need one row for each arc, our dimensions for variable flow selection are the number of arcs also. So this means we need an `identity matrix` for the rows of arcs and columns of arcs.

```{r constraints_1}
Amat <- matrix(0, nrow = nrow(edges), ncol = nrow(edges))
Amat_dir <- rep("<=", nrow(Amat))
Amat_rhs <- c()

for(i in 1:ncol(Amat)){
  Amat[i,i] <- 1
  Amat_rhs <- c(Amat_rhs, edges$capacity[i])
}

Amat
Amat_rhs
Amat_dir
```

The `Node Flow Constraints` are the next to take care of.

For each node, we need to match it's `from` node and `to` node with the appropriate inflow and outflow. If it matches an `inflow` arc, this is increase, so `1` is in the arc column. If it matches an `outflow` arch, this is decrease, so `-1` is in the arc column. Otherwise `0` remains as the placeholder. The sign here is `==` because they must match (i.e., supply = demand). If we require excess at certain points we can set this demand to be higher than zero, but we will not do that here.

```{r constraints_2}
Bmat <- matrix(0, nrow = nrow(nodes), ncol = nrow(edges))
Bmat_dir <- rep("==", nrow(Bmat))
Bmat_rhs <- rep(0, nrow(Bmat))

for(i in 1:nrow(Bmat)){
  node_id <- nodes[i, "id"]
  for(j in 1:ncol(Bmat)){
    edge_from <- edges[j,"from"]
    edge_to <- edges[j, "to"]
    edge_id <- edges[j, "id"]
    
    if(node_id == edge_from){
      Bmat[i,j] = -1
    }
    else if(node_id == edge_to){
      Bmat[i,j] = 1
    }
    else{
      Bmat[i,j] = 0
    }
  }
}

Bmat
Bmat_dir
Bmat_rhs

```

### Objective

Next, the objective is going to be `0` for all values, except our final flow. This we want to maximize.

```{r objective}
f.obj <- c(rep(0, nrow(edges)-1), 1)
f.obj
```

## Solve the LP

Now we unite everything together and drop it into the solver.
Remember, we are trying to maximize flow. So specify `max`.

```{r solve_lp}
f.cons <- rbind(Amat, Bmat)
f.rhs <- c(Amat_rhs, Bmat_rhs)
f.dir <- c(Amat_dir, Bmat_dir)

results <- lp(direction = "max",  
              objective.in = f.obj, 
              const.mat = f.cons, 
              const.dir = f.dir, 
              const.rhs = f.rhs)
results$solution
```
## Visualize Maximum Flow Solution

Since the `results$solution` contain the maximum flow we can push through the pipes as a system, we can add this into our flow component of the `edges` dataframe.

The visual indicates that `2/3` of the capacity was shipped to `node 2`, and `1/2` to `node 1`. After this the next best transfer was from `node 2` to `node 4` maxed out at `2/2`. This is indicated in red. The arcs that have flow are indicated in black, and no flow is indicated by grey. The maximum flow arc is just an `artificial arc` that indicates the maximum flow, this is indicated in purple.

Lastly from the `node 1` to `node 3` `1/4` of the capacity was sent. Then from `node 3` to `node 4` was maxed out at `1/1`. 

Posterior analysis to this model output tells us that sending `2 million tons of oil` from the `source` to `destination 2` and `1 million tons of oil` from the `source` to `destination 1` will push as much flow to `destination 4` as we possibly can.

```{r visual_graph_flow}

# Set the flow
edges$flow <- results$solution

# If the arc is flowing oil, change to black
edges$color[which(edges$flow > 0)] <- "black"

# If the arc is at capacity change it to red
edges$color[which(edges$flow == edges$capacity)] <- "red"

# Last flow is purple
edges$color[nrow(edges)] <- "purple"


g <- igraph::graph_from_data_frame(d = edges,
                                   directed = TRUE,
                                   vertices = nodes)

# Make it look a little more appealing
L = cbind(1:5, 1:5)
CURVE = c(0,0.15, 0.3, 0.45, 0, -0.15, -0.3, 0, 0.15, 0) * 5

# Plot
plot(g,
     layout = L,
     edge.curved = CURVE,
     edge.label = paste("(",E(g)$flow, 
                        ",", 
                        E(g)$capacity, ")", 
                        sep=""), 
     main = "Oil Pipeline Flow | (flow, capacity)", 
     vertex.color = V(g)$color,
     vertex.size = 25,
     vertex.label.cex = 1.6,
     edge.color = E(g)$color,
     edge.width = E(g)$flow*2)

```

### Cleaning Up The Visual

```{r netvis_graph}
visNet_nodes <- nodes %>% 
                dplyr::mutate(label = paste("Location ", id),
                              font.size =10,
                              type = "black",
                              shape = "circle",
                              font = 12)

visNet_edges <- edges %>% 
                dplyr::mutate(label = paste("(",flow,",",capacity,")",sep=""),
                              length = 250,
                              width = flow*2,
                              arrows = "to") 
  
visNetwork(visNet_nodes, 
           visNet_edges,
           width = "100%",
           main = "Maximum Oil Flow | (flow, capacity)")
```



# References

Example taken from the following sources:

<cite>Winston., Wayne. Operations Research, Applications and Algorithms 4th Edition.</cite>